Modal Abstractions for Virtualizing Memory Addresses

### 2023-02-22 programlogic.tex
+

\todo[inline,color=green]{ismail, put the fancy Iris figure for this ghost thing and expand the text above}
\subsection{Address-Space Abstraction}
\label{sec:aspacemodal}
Abstractions we introduce for virtual address spaces are based on two pillars: 1. a global address space invariant asserting the authorative ownership of the ghost map, , summarizing the existence of the page-table walks, and for each physical page address reached by one of these abstract page-table walks we know that there exists concrete physical page-table accesses, i.e. L4-L1 pointsto  2. a modal definition describing the truth of assertions for the resources inside the address space, $[r]P$ indicating that $P$ holds in the virtual address space rooted at $r$.  This is essentially what hybrid logic \ref{} calls a satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here, address space). 
\subsubsection{Global Per-Addressspace Invariant}
\label{sec:peraspaceinvariant}

\subsubsection{Address-Space Modality}
\label{sec:aspacemodalist}
\paragraph{Modal Resource Context}
\label{sec:resourcecontext}
\todo[inline,color=green]{ismail, speak about virtual points-to established under a certain root of an address space}
\paragraph{Interaction}
\label{sec:interaction}
\todo[inline,color=green]{ismail: speak about the interection with ambient logic}
\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.

>>>>

\begin{figure*}
\begin{array}{l}
  \textsf{VirtPtsTo}\stackrel{def}{=} \lambda \tlf\Loc.\\
  \ownGhost\gamma{a}
\caption{Virtual-Pointsto for Sharing Pages}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:virtualpointstosharing}
\end{figure*}
\todo[inline,color=green]{ismail, put the fancy Iris figure for this ghost thing and expand the text above}
\subsection{Address-Space Abstraction}
\label{sec:aspacemodal}
Abstractions we introduce for virtual address spaces are based on two pillars: 1. a global address space invariant asserting the authorative ownership of the ghost map, , summarizing the existence of the page-table walks, and for each physical page address reached by one of these abstract page-table walks we know that there exists concrete physical page-table accesses, i.e. L4-L1 pointsto  2. a modal definition describing the truth of assertions for the resources inside the address space, $[r]P$ indicating that $P$ holds in the virtual address space rooted at $r$.  This is essentially what hybrid logic \ref{} calls a satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here, address space). 
\subsubsection{Global Per-Addressspace Invariant}
\label{sec:peraspaceinvariant}
\begin{figure*}
\begin{array}{l}
  \textsf{ASpaceInv}\stackrel{def}{=} \lambda \tlf\Loc.\\
  \ownGhost\gamma{a}
\caption{Global Address-Space Invarian}
\todo[inline]{These fractions aren't quite right (though I see you added the variance between levels), I can walk you through in our meeting tomorrow.}
  \label{fig:peraspaceinvariant}
\end{figure*}
\subsubsection{Address-Space Modality}
\label{sec:aspacemodalist}
\paragraph{Modal Resource Context}
\label{sec:resourcecontext}
\todo[inline,color=green]{ismail, speak about virtual points-to established under a certain root of an address space}
\paragraph{Interaction}
\label{sec:interaction}
\todo[inline,color=green]{ismail: speak about the interection with ambient logic}
\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3=r_2 \ast r = r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3=r_1 \ast r=r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3=r_1 \ast r=r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.
